---
title: "Financial Model Analysis: Data Loading & Initial Cleaning"
author: "Andrejs Sorstkins"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
# Load necessary tidyverse packages
library(tidyverse)
library(lubridate)
```

## Data Loading and Initial Cleaning (Detailed Explanation)

First, we load the dataset. As defined in our plan, we will treat `;` as the separator and automatically replace any occurrences of the string `"NULL"` with R's standard `NA` format. 

We use `read_delim()` from the `readr` package (part of `tidyverse`), which is faster and more consistent than base R's `read.csv()`.

```{r data_loading}
# Load the data, correctly treating "NULL" as missing values
data <- read_delim("../Data/Assignment_Data.csv", delim = ";", na = c("NULL", "NA", ""), show_col_types = FALSE)

# Take a look at the structure and first few rows
glimpse(data)

cat("Final Dataset Shape: ", dim(data)[1], "rows, ", dim(data)[2], "columns\n")

```

## Type Coercion

Based on the assignment description, we need to explicitly declare which variables are numerical, categorical/ordinal, and dates. This ensures R treats them correctly in later steps.

### Numeric Variables
We will explicitly convert `Variable_1` to `Variable_4`, `Variable_7` to `Variable_11`, and `Variable_15` to `Variable_41` to `numeric` type. We do this in-place using `mutate()` and `across()`.

```{r numeric_coercion}
# Generate a list of numeric column names dynamically to save typing.
# paste0("Variable_", 1:4) creates c("Variable_1", "Variable_2", "Variable_3", "Variable_4")
numeric_vars <- c(
  paste0("Variable_", 1:4),
  paste0("Variable_", 7:11),
  paste0("Variable_", 15:41)
)

# Convert all defined numeric variables to numeric type, overwriting existing columns.
# - mutate() is used to modify the dataframe.
# - across() allows us to apply the as.numeric function to all columns specified in numeric_vars.
# - all_of() ensures we strictly select only the exact names defined in our list.
data <- data %>%
  mutate(across(all_of(numeric_vars), as.numeric))

# Verify numeric conversion
head(data %>% select(all_of(numeric_vars)))
```

### Categorical and Ordinal Variables
These variables should be treated as factors (categories) rather than numeric values or raw strings. We convert them in-place.

```{r categorical_coercion}
# Create a manual list of categorical and ordinal variable names.
# We explicitly list them because they don't follow a simple sequential pattern like the numerics did.
categorical_vars <- c("Variable_5", "Variable_6", "Variable_12", 
                      "Variable_13", "Variable_14", "Variable_45")

# Convert all defined variables to 'factor' type, overwriting existing columns.
# - mutate() modifies the dataframe.
# - across() applies the as.factor function to every column listed in categorical_vars.
# - as.factor() is the R function that converts data into discrete categorical levels.
data <- data %>%
  mutate(across(all_of(categorical_vars), as.factor))

# Verify categorical conversion
head(data %>% select(all_of(categorical_vars)))
```

### Date Unification
The dataset contains multiple date fields potentially in different string formats (some with time, some without). We must unify these into standardized `POSIXct` (datetime) objects using the `lubridate` package.

```{r date_unification}
date_vars <- c("due_date", "first_status_day_date", "first_status_time_of_day", 
               "paid_date", "arrived_date", "Variable_42", "Variable_43", "Variable_44")

# Look at raw date formats before converting
head(data %>% select(all_of(date_vars)))

# Apply parsing to all date variables.
# 1. We use mutate(across(...)) to target all date columns EXCEPT the time-only column.
# 2. lubridate provides parsing functions. dmy_hms() parses Day-Month-Year Hour:Minute:Second.
#    dmy() parses Day-Month-Year with no time.
# 3. coalesce(A, B) is a fallback function. It tries to use A (dmy_hms). If A fails 
#    and returns NA (because there was no time in the string), it falls back to B (dmy).
#    This handles mixed date formats robustly!
data <- data %>%
  mutate(across(c("due_date", "first_status_day_date", "paid_date", 
                  "arrived_date", "Variable_42", "Variable_43", "Variable_44"), 
                ~ coalesce(dmy_hms(., quiet = TRUE), dmy(., quiet = TRUE)))) %>%
  # Handle the time-only column separately using the hms package
  mutate(first_status_time_of_day = hms::as_hms(first_status_time_of_day))

# Verify unified dates
head(data %>% select(all_of(date_vars)))
```

## Handling Missing Data (Optional Step after EDA)

As discovered in our EDA, some columns are too sparse to be useful, while others have massive gaps where the *missingness itself* is a strong signal. We will apply the "Missingness Flag" lifehack here.

```{r missing_data_handling}
# 1. Drop useless columns (>92% missing OR Zero-Variance)
zero_variance_vars <- c("Variable_28", "Variable_3", "Variable_30", 
                        "Variable_31", "Variable_32", "Variable_33", "Variable_34")

data <- data %>% select(-c(Variable_42, Variable_43, Variable_44), 
                        -all_of(zero_variance_vars))

# 2. Create missingness flags for moderate/high missing columns
high_miss_vars <- c("Variable_21", "Variable_38", "Variable_22", "Variable_39", "Variable_20", "Variable_37")

# This creates a new column called 'is_missing_Variable_X' filled with 1s and 0s
data <- data %>%
  mutate(across(all_of(high_miss_vars), ~ as.integer(is.na(.)), .names = "is_missing_{.col}"))

# 3. Safe Median Imputation
# Now we can safely fill the NAs in the original columns with the median
data <- data %>%
  mutate(across(all_of(high_miss_vars), ~ replace_na(., median(., na.rm = TRUE))))

# 4. Drop Redundant (Multicollinear) Features
# We drop these after imputing/flagging to avoid "column not found" errors
redundant_vars <- c("Variable_11", "Variable_26", "Variable_39", "Variable_38", "Variable_37", "Variable_27")
data <- data %>% select(-all_of(redundant_vars))

# Verify
head(data %>% select(starts_with("is_missing_")))
```

## Export Cleaned Data

Now that the initial data loading and type coercion is complete, we export the cleaned dataset to a new CSV file so it can be used for further analysis or modeled in Python.
We use `write_csv()` from the `readr` package because it handles dates and factors consistently and is much faster than base R.

```{r save_data}
# Final Verification
cat("Final Dataset Shape: ", dim(data)[1], "rows, ", dim(data)[2], "columns\n")

# Save the cleaned dataset for modeling in Python
write_csv(data, "../Data/cleaned_assignment_data.csv")
```
